---
title: An R Markdown document converted from "12_EthnicAssoc.ipynb"
output: html_document
---

# Analysis in other ethnicities

```{r}
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(purrr)
library(ggplot2)
library(Matrix, warn.conflicts = FALSE)
library(uwot)
library(survival)
options(repr.plot.res = 300)
```

## Input data

```{r}
recoded_ethn <- read_tsv("../data/recoded_ethn.tsv", show_col_types = FALSE)
```

```{r}
head(recoded_ethn)
```

```{r}
strat_dat <- recoded_ethn %>%
    nest(Data = -c(Ethnicity, sex))
print(strat_dat)
```

```{r}
BIOMARKERS <- c("whr", "sbp", "dbp", "alt", "scr", "crp", "hdl", "tg", "ldl", "fg")
```

```{r}
strat_dat <- strat_dat %>%
    mutate(
        Data = map(
            Data,
            function(DAT){
                DAT %>%
                    mutate(
                        across(
                            all_of(c("bmi", BIOMARKERS)),
                            function(COL){
                                Mean <- mean(COL)
                                SD <- sd(COL)
                                upr <- Mean + (5*SD)
                                lwr <- Mean - (5*SD)
                                COL[COL < lwr] <- NA
                                COL[COL > upr] <- NA
                                return(COL)
                            }
                        )
                    ) %>%
                    drop_na
            }
        )
    )
```

```{r}
print(strat_dat)
```

```{r}
strat_dat %>%
    group_by(Ethnicity) %>%
    summarise(Total = sum(map_dbl(Data, nrow)))
```

## Cluster allocation probabilities and UMAP models

```{r}
load("../data/validclusmod.RData")
```

```{r}
print(validclusmod)
```

```{r}
validclusmod <- validclusmod %>%
    mutate(
        umapmod = map(
            sex, 
            ~load_uwot(paste0("../data/ukb/umapmod_", .x))
        )
    )
print(validclusmod)
```

```{r}
df <- inner_join(validclusmod, strat_dat, by = "sex")
print(df)
```

## Calculation of probabilities

```{r}
df <- df %>%
    mutate(
        resids = map2(
            Data, residmod,
            function(DATA, RESIDMOD){
                sapply(
                    BIOMARKERS,
                    function(biomarker){
                        PREDICTDAT <- as.matrix(cbind(Intercept = 1, DATA[,c("bmi", "age", "smoking")]))
                        MODELDAT <- dplyr::filter(RESIDMOD, Biomarker == biomarker)
                        RESIDSD <- MODELDAT$SDRes
                        COEFDAT <- MODELDAT[,c("Intercept", "bmi", "age", "smoking")]
                        COEFDAT <- t(COEFDAT)
                        PREDVAL <- as.vector(PREDICTDAT %*% COEFDAT)
                        RESIDVAL <- DATA[[biomarker]] - PREDVAL
                        RESIDVAL / RESIDSD
                    }
                )
            }
        )
    )
print(df)
```

```{r}
df <- df %>%
    mutate(
        probs = map2(
            resids, clusmod,
            function(RESIDDF, CLUSMOD){
                mus <- CLUSMOD$validclus_centers
                mus <- map(mus, function(mu){ mu[BIOMARKERS] })
                covmats <- CLUSMOD$validclus_covmats
                covmats <- map(covmats, function(covmat){ covmat[BIOMARKERS, BIOMARKERS] })
                wts <- CLUSMOD$validclus_weights
                pdfs <- mapply(
                    function(mu, covmat){ mvtnorm::dmvnorm(RESIDDF, mu, covmat) },
                    mus, covmats
                )
                L <- pdfs %*% diag(wts)
                colnames(L) <- paste0("prob", CLUSMOD$validclus_name)
                tibble(data.frame(L / rowSums(L)))
            }
        )
    )
print(df)
```

```{r}
head(df$probs[[1]])
```

```{r}
cluscolmap <- validclusmod %>%
    transmute(
        clusmod = map(clusmod, select, Cluster = validclus_name)
    ) %>%
    unnest(clusmod) %>%
    unique %>%
    mutate(ClusCol = scales::hue_pal()(n()))
cluscolmap
```

```{r}
PercDat <- df %>%
    transmute(
        sex, Ethnicity,
        probs = map(probs, colSums),
        probs = map(
            probs, 
            ~tibble(
                Cluster = gsub("prob", "", names(.x)),
                N = .x,
                Perc = .x/sum(.x)
            )
        )
    ) %>%
    unnest(probs)
head(PercDat)
```

```{r}
PercDatRef <- validclusmod %>%
    transmute(
        sex, 
        Ethnicity = "European",
        clusmod = map(
            clusmod, select, 
            Cluster = validclus_name, 
            Perc = validclus_weights
        )
    ) %>%
    unnest(clusmod)
head(PercDatRef)
```

```{r}
ClusOrd <- PercDatRef %>%
    group_by(sex) %>%
    arrange(desc(Perc), .by_group = TRUE) %>%
    mutate(ordn = row_number()) %>%
    ungroup %>%
    select(sex, Cluster, ordn)
ClusOrd
```

```{r}
options(repr.plot.height = 4, repr.plot.width = 12)
PercDat %>%
    select(-N) %>%
    bind_rows(PercDatRef) %>%
    inner_join(cluscolmap, by = "Cluster") %>%
    mutate(
        Perc2 = ifelse(Perc > 0.125, Perc, NA)
    ) %>%
    pivot_longer(c(Perc, Perc2), names_to = "PercLevel", values_to = "PercValue") %>%
    mutate(
        Base = case_when(
            PercLevel == "Perc" ~ 0,
            PercLevel == "Perc2" & !is.na(PercValue) ~ .6
        ),
        PercValue = ifelse(PercLevel == "Perc" & PercValue > .125, .125, PercValue),
        PercLevel = factor(PercLevel, levels = c("Perc2", "Perc")),
        across(c(Base, PercValue), \(x) 100 * x),
        Ethnicity = factor(Ethnicity, levels = c("European", "African", "SouthAsian"))
    ) %>%
    inner_join(ClusOrd, by = c("sex", "Cluster")) %>%
    ggplot(aes(reorder(Cluster, ordn), PercValue)) +
    geom_rect(
        aes(
            xmin = Cluster, xmax = Cluster,
            ymin = Base, ymax = PercValue,
            color = ClusCol
        ),
        linewidth = 8, na.rm = TRUE
    ) +
    scale_color_identity() +
    ggh4x::facet_nested(
        PercLevel ~ sex + Ethnicity,
        scales = "free", space = "free_x"
    ) +
    ggh4x::force_panelsizes(
        rows = c(1, 3)
    ) +
    ggh4x::scale_y_facet(
        PercLevel == "Perc",
        breaks = c(0, 5, 10),
        expand = c(.05, 0, 0, 0) 
    ) +
    ggh4x::scale_y_facet(
        PercLevel == "Perc2",
        breaks = c(60,80),
        limits = c(60,90),
        expand = expansion(mult = c(0, NA))
    ) +
    theme_bw() +
    theme(strip.text.y = element_blank()) +
    labs(x = NULL, y = "%")
```

## Comparing proportions

```{r}
RefDat <- validclusmod %>%
    transmute(
        sex, 
        residtab = map(
            umapmod,
            ~map(
                1:.x$nn_index$ann$getNItems(),
                function(i){
                    .x$nn_index$ann$getItemsVector(i-1)
                }
            )
        ),
        resids = map(residtab, do.call, what = rbind),
        probs = map2(
            resids, clusmod,
            function(RESIDDF, CLUSMOD){
                pdfs <- mapply(
                    function(mu, covmat){ mvtnorm::dmvnorm(RESIDDF, mu, covmat) },
                    CLUSMOD$validclus_centers, 
                    CLUSMOD$validclus_covmats
                )
                L <- pdfs %*% diag(CLUSMOD$validclus_weights)
                colnames(L) <- paste0("prob", CLUSMOD$validclus_name)
                tibble(data.frame(L / rowSums(L)))
            }
        )
    )
print(RefDat)
```

```{r}
RefN <- RefDat %>%
    transmute(
        sex, Ethnicity = "European",
        probs = map(probs, colSums),
        probs = map(
            probs, 
            ~tibble(
                Cluster = gsub("prob", "", names(.x)),
                N = .x, Total = sum(.x)
            )
        )
    ) %>%
    unnest(probs)
RefN
```

```{r}
RefNList <- nest(RefN, RefD = -c(sex, Cluster))
print(RefNList)
```

```{r}
BinomDat <- PercDat %>%
    select(-Perc) %>%
    group_by(sex, Ethnicity) %>%
    mutate(Total = sum(N)) %>%
    ungroup %>%
    nest(Data = -c(sex, Cluster, Ethnicity)) %>%
    inner_join(RefNList, by = join_by(sex, Cluster)) %>%
    mutate(
        Data = map2(Data, Ethnicity, ~mutate(.x, Ethnicity = .y)),
        Data = map2(RefD, Data, bind_rows),
        RefD = NULL,
        Data = map(
            Data, mutate, 
            Nnot = Total - N,
            Ethnicity = relevel(factor(Ethnicity), ref = "European")
        )
    )
print(BinomDat)
```

```{r}
BinomDat$Data[[1]]
```

```{r}
BinomDat <- BinomDat %>%
    mutate(
        Mod = map(Data, ~glm(cbind(round(N), round(Nnot)) ~ Ethnicity, data = .x, family = binomial))
    )
print(BinomDat)
```

```{r}
BinomRes <- BinomDat %>%
    transmute(
        sex, Cluster, Ethnicity,
        Data = map(Data, filter, Ethnicity != "European"),
        Data = map(Data, select, -Ethnicity),
        Mod = map(Mod, broom::tidy, exponentiate = TRUE, conf.int = TRUE),
        Mod = map(Mod, filter, term != "(Intercept)"),
        Mod = map(Mod, select, -c(term, std.error, statistic))
    ) %>%
    unnest(c(Data, Mod)) %>%
    mutate(
        padj = p.adjust(p.value, "fdr")
    )
BinomRes
```

```{r}
write_tsv(BinomRes, "../data/BinomRes_ethn.tsv")
```

```{r}
SigBinomRes <- BinomRes %>% 
    filter(padj < 0.05)
SigBinomRes
```

```{r}
SigUpBinomRes <- SigBinomRes %>% 
    filter(estimate > 1) %>%
    mutate(across(c(estimate, conf.low, conf.high), ~round(.x, digits = 2))) %>%
    arrange(Ethnicity, sex, desc(estimate))
SigUpBinomRes
```

## UMAP projection

Now we can plot the deviations from BMI of the new data points on top of the UK biobank projection:

```{r}
newcoords <- df %>%
    transmute(
        sex, Ethnicity,
        umapg = map2(
            resids, umapmod, 
            function(newx, mod){
                data.frame(umap_transform(newx, mod))
            }
        )
    ) %>%
    unnest(umapg)
head(newcoords)
```

```{r}
basecoords <- df %>%
    transmute(
        sex, Ethnicity,
        umapg = map(umapmod, ~data.frame(.x$embedding))
    ) %>%
    unnest(umapg)
head(basecoords)
```

```{r}
options(repr.plot.height = 5, repr.plot.width = 5)
embed <- basecoords %>%
    ggplot(aes(X1, X2)) +
    geom_point(shape = ".", alpha = .1) +
    geom_point(data = data.frame(newcoords),
               color = "red", shape = ".") +
    facet_grid(Ethnicity~sex) +
    theme_bw() +
    labs(x = "UMAP1", y = "UMAP2")
embed
```

## Covariate data

```{r}
covar_dat <- readr::read_tsv("../data/covar_dat.tsv", show_col_types = FALSE)
head(covar_dat)
```

## Prevalence

```{r}
df <- df %>%
    mutate(
        # Joining complete biomarker data with allocation probabilities
        joindat = map2(Data, probs, bind_cols),
        # Adding covariate data
        joindat = map(joindat, inner_join, covar_dat, by = "eid")
    )
print(df)
```

```{r}
prevdat <- df %>%
    transmute(
        sex, Ethnicity,
        propdat = map(
            joindat,
            function(DAT){
                tibble(
                    Cluster = names(DAT)[grepl("^prob", names(DAT))]
                ) %>%
                    mutate(
                        clusprobdat = map(
                            Cluster,
                            function(CL){
                                map(
                                    c("HT", "CHD", "Stroke", "LiverFailure", "RA", "T2D", "T1D", 
                                      "Insulin", "AntiDM", "AntiHT", "LipidLower"),
                                    function(DX){
                                        tibble(
                                            Disease = DX,
                                            N = sum(DAT[[CL]]),
                                            Ncases = sum(DAT[[DX]] * DAT[[CL]])
                                        )
                                    }
                                ) %>%
                                    bind_rows
                            }
                        )
                    ) %>%
                    unnest(clusprobdat)
            }
        )
    ) %>%
    unnest(propdat) %>%
    mutate(Cluster = gsub("prob", "", Cluster))
```

```{r}
head(prevdat)
```

### Odds ratios relative to the concordant cluster

```{r}
ORdat <- prevdat %>%
    nest(Data = -c(sex, Ethnicity, Disease)) %>%
    mutate(
        Data = map(
            Data,
            mutate,
            Cluster = factor(Cluster),
            Cluster = relevel(Cluster, ref = "BC")
        ),
        mod = map(
            Data,
            ~glm(cbind(round(Ncases), round(N - Ncases)) ~ Cluster, data = .x, family = binomial)
        ),
        modres = map(mod, broom::tidy),
        modres = map(modres, filter, term != "(Intercept)"),
        modres = map(modres, mutate, Cluster = gsub("Cluster", "", term), term = NULL),
        modres = map2(Data, modres, left_join, join_by(Cluster))
    ) %>%
    select(sex, Ethnicity, Disease, modres) %>%
    unnest(modres) %>%
    mutate(
        OR = exp(estimate),
        lwr = exp(estimate - qnorm(1 - .05/2) * std.error),
        upr = exp(estimate + qnorm(1 - .05/2) * std.error)
    ) %>%
    select(-c(estimate, std.error, statistic))
head(ORdat)
```

```{r}
DiseaseDat <- ORdat %>%
    filter(Disease %in% c("HT", "CHD", "Stroke", "LiverFailure", "RA", "T2D", "T1D")) %>%
    mutate(padj = p.adjust(p.value, "fdr"))
head(DiseaseDat)
```

```{r}
write_tsv(DiseaseDat, "../data/DiseaseDat_ethn.tsv")
```

```{r}
DiseaseDat %>%
    filter(padj < 0.01)
```

```{r}
options(repr.plot.width = 10, repr.plot.height = 5)
DiseaseDat %>%
    drop_na %>%
    ggplot(aes(OR, Cluster)) +
    geom_vline(xintercept = 1, lty = "dashed") +
    geom_linerange(aes(xmin = lwr, xmax = upr)) +
    geom_point() +
    ggh4x::facet_nested(sex + Ethnicity ~ Disease, scales = "free", space = "free_y") +
    labs(x = "OR relative to BC", y = NULL, color = "Model") +
    theme_bw() +
    theme(
        axis.text.x = element_text(size = 5)
    )
```

```{r}
MedDat <- ORdat %>%
    filter(Disease %in% c("Insulin", "AntiHT", "LipidLower")) %>%
    mutate(padj = p.adjust(p.value, "fdr"))
head(MedDat)
```

```{r}
write_tsv(MedDat, "../data/MedDat_ethn.tsv")
```

```{r}
MedDat %>%
    filter(padj < 0.01)
```

```{r}
options(repr.plot.width = 6, repr.plot.height = 5)
ORdat %>%
    drop_na %>%
    filter(Disease %in% c("Insulin", "AntiHT", "LipidLower")) %>%
    ggplot(aes(OR, Cluster)) +
    geom_vline(xintercept = 1, lty = "dashed") +
    geom_linerange(aes(xmin = lwr, xmax = upr)) +
    geom_point() +
    ggh4x::facet_nested(sex + Ethnicity ~ Disease, scales = "free", space = "free_y") +
    labs(x = "OR relative to BC", y = NULL, color = "Model") +
    theme_bw() +
    theme(
        axis.text.x = element_text(size = 5)
    )
```

## Incidence

```{r}
df <- df %>%
    expand_grid(
        Outcome = c("mace", "dm")
    )
print(df)
```

```{r}
df <- df %>%
    mutate(
        SurvDat = map(
            Outcome,
            ~readr::read_tsv(paste0("../data/surv", .x, "dat_ethn.tsv"), show_col_types = FALSE)
        ),
        SurvDat = map2(
            SurvDat, joindat, inner_join, by = "eid"
        ),
        SurvDat = map2(
            Outcome, SurvDat,
            ~switch(
                .x,
                mace = filter(.y, CHD == 0, Stroke == 0, PAD == 0),
                dm = filter(.y, T2D == 0, T1D == 0, Insulin == 0, AntiDM == 0)
            )
        ),
        SurvDat = map(SurvDat, select, -where(\(x) all(x == 0)))
    )
print(df)
```

## Data for modelling

```{r}
egfrcalc <- function(scr, sex, age){
    scr <- scr / 88.42
    alpha <- ((sex == "Female") * -0.241) + ((sex == "Male") * -0.302)
    kappa <- ((sex == "Female") * 0.7) + ((sex == "Male") * 0.9)
    creat_kappa <- scr / kappa
    minkappa <- pmin(creat_kappa, 1) 
    maxkappa <- pmax(creat_kappa, 1)
    minkappa_alpha <- minkappa^alpha
    maxkappa_exp <- maxkappa^(-1.200)
    age_term <- 0.9938^age
    sex_term <- ((sex == "Female") * 1.012) + ((sex == "Male") * 1) 
    142 * minkappa_alpha * maxkappa_exp * age_term * sex_term
}
```

```{r}
df <- df %>%
    mutate(
        SurvRegDat = pmap(
            list(SurvDat, Outcome, sex),
            function(DAT, OUT, SEX){
                alrprobdat <- DAT %>%
                    transmute(
                        across(starts_with("prob"), \(x) ifelse(x < 1e-15, 1e-15, x)),
                        across(starts_with("prob"), \(x) ifelse(x > (1 - 1e-15), (1 - 1e-15), x)),
                        across(starts_with("prob"), \(x) log(x/probBC)),
                        probBC = NULL
                    )
                names(alrprobdat) <- paste0("alr", names(alrprobdat))
                switch(
                    OUT,
                    mace = transmute(
                        DAT,
                        outcome_value, outcome_timeyrs,
                        age, smoking, bmi,
                        sbp,
                        tchol = hdl + ldl + (tg/2.2),
                        hdl,
                        fg,
                        whr, alt, crp,
                        T2D, T2Dage,
                        lnegfr = log(egfrcalc(scr, SEX, age)),
                        lnegfrsq = lnegfr^2,
                        age_smoking = age*smoking,
                        age_sbp = age*sbp,
                        age_tchol = age*tchol,
                        age_hdl = age*hdl,
                        age_fg = age*fg,
                        age_lnegfr = age*lnegfr,
                        age_t2d = age*T2D,
                        Insulin, AntiHT, LipidLower
                    ),
                    dm = transmute(
                        DAT,
                        outcome_value, outcome_timeyrs,
                        age, smoking, bmi,
                        whr, sbp, dbp, alt, scr, crp, hdl, tg, ldl, fg,
                        CHD, Stroke, AntiHT, LipidLower
                    )
                ) %>%
                     bind_cols(alrprobdat)   
            }
        )
    )
print(df)
```

## Survival models

```{r}
df <- df %>%
    mutate(
        NullMod = map(
            SurvRegDat, 
            ~coxph(Surv(time = outcome_timeyrs, event = outcome_value) ~ 1, data = .x)
        ),
        ModBase = map(
            SurvRegDat,
            ~{
                .x <- select(.x, -starts_with("alr"))
                coxph(Surv(time = outcome_timeyrs, event = outcome_value) ~ ., data = .x)
            }
        ),
        ModClus = map(
            SurvRegDat, 
            ~coxph(Surv(time = outcome_timeyrs, event = outcome_value) ~ ., data = .x)
        )
    )
print(df)
```

## Added value

```{r}
addedvaltab <- df %>%
    transmute(
        Outcome, sex, Ethnicity,
        LL0 = map_dbl(NullMod, logLik),
        LLBase = map_dbl(ModBase, logLik),
        NVBase = map_dbl(ModBase, \(MOD) sum(!is.na(MOD$coefficients))),
        LLClus = map_dbl(ModClus, logLik),
        NVClus = map_dbl(ModClus, \(MOD) sum(!is.na(MOD$coefficients))),
        LRTstat = 2 * abs(LLClus - LLBase),
        LRTdf = NVClus - NVBase,
        LRTp = stats::pchisq(
            q = LRTstat,
            df = LRTdf,
            lower.tail = FALSE
        ),
        LRTpadj = p.adjust(LRTp, "fdr"),
        AdeqInd = (LLBase - LL0) / (LLClus - LL0),
        CBase = purrr::map_dbl(ModBase, \(MOD) MOD[["concordance"]][["concordance"]]),
        CseBase = purrr::map_dbl(ModBase, \(MOD) MOD[["concordance"]][["std"]]),
        CClus = purrr::map_dbl(ModClus, \(MOD) MOD[["concordance"]][["concordance"]]),
        CseClus = purrr::map_dbl(ModClus, \(MOD) MOD[["concordance"]][["std"]]),
        Cdat = purrr::map2(ModClus, ModBase, survival::concordance),
        Cdiff = purrr::map_dbl(Cdat, \(MOD) c(1, -1) %*% coef(MOD)),
        Cdiffse = purrr::map_dbl(Cdat, \(MOD) sqrt(c(1, -1) %*% vcov(MOD) %*% c(1, -1))),
        Cdiffp = 2*pnorm(-abs(Cdiff / Cdiffse)),
        Cdiffpadj = p.adjust(Cdiffp, "fdr"),
        Cdat = NULL
    )
addedvaltab
```

```{r}
write_tsv(addedvaltab, "../data/addedvaltab_ethn.tsv")
```

```{r}
addedvaltab %>%
    select(Outcome, sex, Ethnicity, LRTp, Cdiffp) %>%
    filter(LRTp < 0.05)
```

#### Effect of increasing 10% in a discordant cluster at the expense of the concordant cluster on outcomes

```{r}
shift10pc <- validclusmod %>%
    expand_grid(Outcome = c("mace", "dm")) %>%
    transmute(
        sex, Outcome,
        NewD = map2(
            clusmod, residmod,
            function(CLUSMOD, RESIDMOD){
                averesidshift <- CLUSMOD %>%
                    expand_grid(
                        Cluster = CLUSMOD$validclus_name
                    ) %>%
                    mutate(
                        # Producing the shifts
                        toadd = case_when(
                            Cluster == "BC" ~ 0,
                            Cluster == validclus_name ~ .1,
                            validclus_name == "BC" ~ -.1,
                            TRUE ~ 0
                        ),
                        validclus_weights = validclus_weights + toadd
                    ) %>%
                    group_by(Cluster) %>%
                    summarise(
                        # Generating data with the shift
                        RGD = list(
                            MGMM::rGMM(
                                n = 5e5, 
                                d = length(validclus_centers[[1]]),
                                k = n(),
                                pi = validclus_weights,
                                means = validclus_centers, 
                                covs = validclus_covmats
                            ) %>%
                                colMeans %>%
                                setNames(
                                    names(validclus_centers[[1]])
                                )
                        )
                    ) %>%
                    mutate(
                        # Values corresponding to the shift
                        predbmvals = map(
                            RGD,
                            ~tibble(
                                Biomarker = names(.x), sdresidval = .x
                            ) %>%
                                inner_join(RESIDMOD, by = "Biomarker") %>%
                                mutate(
                                    sdresid = sdresidval * SDRes,
                                    bmi = 30 * bmi,
                                    age = 60 * age,
                                    bmvalue = rowSums(pick(Intercept, bmi, age, sdresid))
                                ) %>%
                                {setNames(.$bmvalue, .$Biomarker)} %>%
                                c(bmi = 30, age = 60) %>% t
                        ),
                        # Calculating probabilities and log-ratios
                        predprobs = map(
                            RGD, 
                            ~{
                                pdfs <- mapply(
                                    function(mu, covmat){ 
                                        mvtnorm::dmvnorm(
                                            .x, mu, covmat
                                        ) 
                                    },
                                    CLUSMOD$validclus_centers, 
                                    CLUSMOD$validclus_covmats
                                )
                                L <- pdfs %*% diag(CLUSMOD$validclus_weights)
                                colnames(L) <- CLUSMOD$validclus_name
                                L / rowSums(L)
                            }
                        ),
                        alrprobs = map(
                            predprobs,
                            ~{
                                cclus <- colnames(.x) == "BC"
                                alrp <- .x[,!cclus]/.x[,cclus]
                                names(alrp) <- paste0("alrprob", names(alrp))
                                t(alrp)
                            }
                        ),
                        joindat = map2(predbmvals, alrprobs, cbind)
                    ) %>%
                    transmute(Cluster, joindat = map(joindat, data.frame)) %>%
                    unnest(joindat)
            }
        )
    )
print(shift10pc)
```

```{r}
shift10pc$NewD[[1]]
```

```{r}
shift10pc <- shift10pc %>%
    mutate(
        NewD = pmap(
            list(Outcome, NewD, sex),
            function(OUT, DAT, SEX){
                switch(
                    OUT,
                    mace = transmute(
                        DAT,
                        Cluster,
                        sbp,
                        tchol = hdl + ldl + (tg/2.2),
                        hdl,
                        fg,
                        whr, alt, crp,
                        lnegfr = log(egfrcalc(scr, SEX, age)),
                        lnegfrsq = lnegfr^2,
                        age_sbp = age*sbp,
                        age_tchol = age*tchol,
                        age_hdl = age*hdl,
                        age_fg = age*fg,
                        age_lnegfr = age*lnegfr,
                        across(starts_with("alrprob"))
                    ),
                    dm = transmute(
                        DAT,
                        Cluster,
                        whr, sbp, dbp, alt, scr, crp, hdl, tg, ldl, fg,
                        across(starts_with("alrprob"))
                    )
                )
            }
        )
    )
print(shift10pc)
```

```{r}
shift10pc <- shift10pc %>%
    mutate(
        NewD = map(
            NewD,
            ~{
                mat <- as.matrix(select(.x, -Cluster))
                rownames(mat) <- .x$Cluster
                cclus <- rownames(mat) == "BC"
                t(t(mat[!cclus,]) - mat[cclus,])
            }
        )
    )
print(shift10pc)
```

```{r}
shift10pc$NewD[[1]]
```

```{r}
df <- df %>%
    inner_join(
        shift10pc, by = join_by(sex, Outcome)
    )
print(df)
```

And now we can calculate the effect of each shift:

```{r}
shift10est <- df %>%
    transmute(
        sex, Ethnicity, Outcome,
        ModClus = map2(
            ModClus, NewD,
            function(MOD, D){
                BETAS <- coef(MOD)
                nm <- intersect(colnames(D), names(BETAS))
                covm <- vcov(MOD)[nm,nm]
                data.frame(
                    Cluster = gsub("alrprob", "", rownames(D)),
                    estimate = drop(D[,nm] %*% BETAS[nm]),
                    se = apply(
                        D[,nm], 1,
                        \(clus){ 
                            sqrt(drop(clus %*% covm %*% clus))
                        }
                    )
                )
            }
        )
    ) %>%
    unnest(ModClus) %>%
    mutate(
        pval = 2 * pnorm(-abs(estimate/se)),
        HR = exp(estimate),
        lwr = exp(estimate - qnorm(.975)*se),
        upr = exp(estimate + qnorm(.975)*se),
        padj = p.adjust(pval, "fdr")
    )
head(shift10est)
```

```{r}
write_tsv(shift10est, "../data/shift10est_ethn.tsv")
```

```{r}
shift10est %>%
    filter(padj < 0.05)
```

```{r}
options(repr.plot.width = 4, repr.plot.height = 5)
shift10est %>%
    ggplot(aes(HR, Cluster)) +
    geom_vline(xintercept = 1, lty = "dashed") +
    geom_linerange(aes(xmin = lwr, xmax = upr)) +
    geom_point() +
    ggh4x::facet_nested(sex + Ethnicity ~ Outcome, scales = "free", space = "free_y") +
    labs(y = NULL) +
    theme_bw()
```

