---
title: An R Markdown document converted from "00_DataPrep.ipynb"
output: html_document
---

# Data preparation

---

## Libraries

```{r}
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(purrr)
```

---

## Baseline data

```{r}
tab_name <- "~/projects/DVA/Data/UKBiobank/ukb_57232/data_files/41806.tab"
```

Locating columns needed:

```{r}
tab_vars <- tribble(
    ~colid, ~colnam, ~coltyp,
    ## Basic
    "f.eid", "eid", "n",
    "f.53.0.0", "date0", "D",
    "f.31.0.0", "sex", "n",
    "f.21003.0.0", "age", "n",
    ## Anthropometric
    "f.21001.0.0", "bmi", "n",
    "f.48.0.0", "waist_circumference", "n",
    "f.49.0.0", "hip_circumference", "n",
    "f.50.0.0", "height", "n",
    ## Blood pressure
    "f.4080.0.0", "sbp_01", "n",
    "f.4080.0.1", "sbp_02", "n",
    "f.4079.0.0", "dbp_01", "n",
    "f.4079.0.1", "dbp_02", "n",
    ## Liver function
    "f.30620.0.0", "alt", "n",
    ## Kidney function
    "f.30700.0.0", "creatinine", "n",
    ## Inflammatory markers
    "f.30710.0.0", "crp", "n",
    ## Lipid fractions
    "f.30760.0.0", "hdl", "n",
    "f.30780.0.0", "ldl", "n",
    "f.30870.0.0", "tg", "n",
    ## Glycemia
    "f.30740.0.0", "glucose", "n",
    ## Smoking
    "f.1239.0.0", "current_smoker", "n",
    ## Caucasian ancestry
    "f.22006.0.0", "caucasian", "n"
)
```

Collecting positions of these columns in the data:

```{r}
tab_vars <- paste("head -1", tab_name, "| tr '\t' '\n'") %>%
    pipe %>%
    read_table(col_names = "colid", col_types = "c") %>%
    ## Position of columns in table
    mutate(colpos = row_number()) %>%
    ## Extracting position and name of variables of interest
    inner_join(tab_vars, by = "colid")
head(tab_vars)
```

Fasting time (preprocessed from a second application):

```{r}
fast_time <- read_tsv("~/projects/DVA/Data/UKBiobank/fasting_time_ukb.tsv", show_col_types = FALSE)
head(fast_time)
```

Preparing to join with rest of columns:

```{r}
fast_time <- transmute(fast_time, eid = eid_pascal, fasting_time)
head(fast_time)
```

Preparing the `bash` command to extract columns of interest:

```{r}
column_cmd <- paste("cut -f", paste(tab_vars$colpos, collapse = ","), tab_name)
cat(column_cmd)
```

Executing the command and passing to `R`:

```{r}
crude_dat <- read_tsv(
    file = pipe(column_cmd),
    ## Replacing column header with variable names
    skip = 1, col_names = tab_vars$colnam,
    ## All columns are numeric
    col_types = paste(tab_vars$coltyp, collapse = "")
) %>%
    ## Adding fasting time
    inner_join(fast_time, by = "eid")
head(crude_dat)
```

Recoding data:

```{r}
recoded_dat <- crude_dat %>%
    ## Only European descent individuals
    filter(caucasian == 1) %>%
    select(-caucasian) %>%
    transmute(eid, age, 
              sex = ifelse(sex == 0, "Female", "Male"),
              ## Anthropometric
              bmi, 
              whr = waist_circumference / hip_circumference,
              ## Blood pressure
              sbp = rowMeans(cbind(sbp_01, sbp_02), na.rm = TRUE),
              dbp = rowMeans(cbind(dbp_01, dbp_02), na.rm = TRUE),
              ## Liver function
              alt,
              ## Kidney function
              scr = creatinine,
              ## Inflammatory markers
              crp,
              ## Lipid fractions
              hdl, tg, ldl,
              ## Glycemia - At least fasting time of 4 hours
              fg = ifelse(fasting_time >= 4, glucose, NaN),
              ## Current smoking status
              smoking = case_when(current_smoker %in% c(1,2) ~ 1,
                                  current_smoker == 0 ~ 0,
                                  TRUE ~ NaN)) %>%
    ## Only complete cases
    drop_na
head(recoded_dat)
nrow(recoded_dat)
```

---

## Pre-existing conditions

```{r}
tab_vars <- tibble(
    colid = c("f.eid",  "f.31.0.0"),
    colnam = c("eid", "sex"),
) %>%
    bind_rows(
        data.frame(colid = paste0("f.20002.0.", 0:33),
                   colnam = paste("noncancercode", 1:34, sep = "_")),
        data.frame(colid = paste0("f.20009.0.", 0:33),
                   colnam = paste("noncancerage", 1:34, sep = "_"))
    )
tab_vars <- paste("head -1", tab_name, "| tr '\t' '\n'") %>%
    pipe %>%
    read_table(col_names = "colid", col_types = "c") %>%
    ## Position of columns in table
    mutate(colpos = row_number()) %>%
    ## Extracting position and name of variables of interest
    inner_join(tab_vars, by = "colid")
head(tab_vars)
```

```{r}
column_cmd <- paste("cut -f", paste(tab_vars$colpos, collapse = ","), tab_name)
cat(column_cmd)
```

```{r}
dx_dat <- read_tsv(
    file = pipe(column_cmd),
    ## Replacing column header with variable names
    skip = 1, col_names = tab_vars$colnam,
    ## All columns are numeric
    col_types = cols(.default = "n")
)
head(dx_dat)
```

```{r}
nrow(dx_dat)
```

Long format

```{r}
dx_dat2 <- dx_dat %>%
    pivot_longer(
        -c(eid, sex),
        names_sep = "_",
        names_to = c(".value", NA),
        values_drop_na = TRUE
    )
head(dx_dat2)
```

Joining with coding:

```{r}
dx_coding <- read_tsv("~/projects/DVA/Data/UKBiobank/ukb_57232/codings/coding6.tsv",
                      show_col_types = FALSE)
head(dx_coding)
```

```{r}
dx_dat2 <- dx_dat2 %>%
    inner_join(dx_coding[, c("coding", "meaning")],
               by = c("noncancercode" = "coding"))
head(dx_dat2)
```

### Selection of pre-existing conditions

```{r}
selected_dx <- dx_dat2 %>%
    group_by(eid) %>%
    summarise(
        HT = any(meaning %in% c("hypertension", 
                                "essential hypertension")),
        CHD = any(meaning %in% c("heart/cardiac problem",
                                 "angina", "heart attack/myocardial infarction")),
        Stroke = any(meaning %in% c("stroke", "ischaemic stroke",
                                    "transient ischaemic attack (tia)",
                                    "subarachnoid haemorrhage")),
        PAD = any(meaning %in% c("arterial embolism", "peripheral vascular disease",
                                 "leg claudication/ intermittent claudication")),
        CKD = any(meaning %in% c("renal/kidney failure",
                                 "renal failure requiring dialysis",
                                 "renal failure not requiring dialysis")),
        LiverFailure = any(meaning %in% c("liver failure/cirrhosis",
                                          "primary biliary cirrhosis",
                                          "alcoholic liver disease / alcoholic cirrhosis")),
        RA = any(meaning == "rheumatoid arthritis"),
        T2D = any(meaning %in% c("diabetes", "type 2 diabetes", "diabetic eye disease", 
                                 "diabetic neuropathy/ulcers")),
        T1D = any(meaning %in% c("type 1 diabetes")),
        T2Dage = noncancerage[
            meaning %in% c("diabetes", "type 2 diabetes", "diabetic eye disease", 
                           "diabetic neuropathy/ulcers")
        ][1]
    ) %>%
    mutate(across(-c(eid, T2Dage), ~.x * 1)) %>%
    right_join(dx_dat[,"eid"], by = "eid") %>%
    mutate(
        across(-eid, ~replace_na(.x, 0)),
        T2Dage = ifelse(T2Dage < 0, NaN, T2Dage),
        T1D = ifelse(T2D == 1 & T2Dage < 35, 1, T1D),
        T2D = ifelse(T1D == 1, 0, T2D),
        T2Dage = ifelse(T2D == 1 & T1D == 0, T2Dage, 0)
    ) %>%
    drop_na
head(selected_dx)
nrow(selected_dx)
```

---

## Medication data

```{r}
tab_vars <- tibble(
    colid = "f.eid",
    colnam = "eid",
) %>%
    bind_rows(
        data.frame(colid = paste0("f.20003.0.", 0:47),
                   colnam = paste("medcode", 1:48, sep = "_"))
    )
tab_vars <- paste("head -1", tab_name, "| tr '\t' '\n'") %>%
    pipe %>%
    read_table(col_names = "colid", col_types = "c") %>%
    ## Position of columns in table
    mutate(colpos = row_number()) %>%
    ## Extracting position and name of variables of interest
    inner_join(tab_vars, by = "colid")
head(tab_vars)
```

```{r}
column_cmd <- paste("cut -f", paste(tab_vars$colpos, collapse = ","), tab_name)
cat(column_cmd)
```

```{r}
meds_dat <- read_tsv(
    file = pipe(column_cmd),
    ## Replacing column header with variable names
    skip = 1, col_names = tab_vars$colnam,
    ## All columns are numeric
    col_types = cols(.default = "n")
)
head(meds_dat)
```

Long format

```{r}
meds_dat2 <- meds_dat %>%
    pivot_longer(
        -eid,
        names_sep = "_",
        names_to = c(".value", NA),
        values_drop_na = TRUE
    )
head(meds_dat2)
```

Joining with coding:

```{r}
meds_coding <- read_tsv("~/projects/DVA/Data/UKBiobank/ukb_57232/codings/coding4.tsv",
                      show_col_types = FALSE)
head(meds_coding)
```

```{r}
meds_dat2 <- meds_dat2 %>%
    inner_join(meds_coding,
               by = c("medcode" = "coding"))
head(meds_dat2)
```

Map of UK Biobank medication to ATC codes (available for some medications only):

```{r}
ATC_matches <- read_tsv("~/projects/DVA/Data/UKBiobank/ukb_57232/codings/ATC_matches.tsv", 
                        col_types = "ncc")
head(ATC_matches)
```

Selected codes to use:

- A10A - Insulin
- A10B - Diabetes drugs
- C01 - Cardiac therapy
- C02 - Antihypertensives
- C03 - Diuretics
- C07 - Beta blocking agents
- C08 - Calcium channel blockers
- C09 - RA system drugs
- C10 - Lipid lowering

```{r}
ATC_matches_sel <- ATC_matches %>%
    filter(grepl("^A10|^C01|^C02|^C03|^C07|^C08|^C09|^C10", ATC_codes)) %>%
    unique
head(ATC_matches_sel)
```

```{r}
selmeds <- meds_dat2 %>%
    inner_join(ATC_matches_sel, by = c("medcode" = "ukb_code", 
                                       "meaning" = "med_name"))
head(selmeds)
```

Ocurrence of selected medication:

```{r}
selected_meds <- selmeds %>%
    group_by(eid) %>%
    summarise(
        Insulin = any(ATC_codes %in% c("A10A")),
        AntiDM = any(ATC_codes %in% c("A10B")),
        AntiHT = any(grepl("^C01|^C02|^C03|^C07|^C08|^C09", ATC_codes)),
        LipidLower = any(grepl("^C10", ATC_codes))
    ) %>%
    mutate(across(-eid, ~.x * 1)) %>%
    right_join(meds_dat[,"eid"], by = "eid") %>%
    mutate(across(-eid, ~replace_na(.x, 0)))
head(selected_meds)
nrow(selected_meds)
```

---

## Joining pre-existing conditions and medication data

```{r}
covar_dat <- inner_join(selected_dx, selected_meds, by = "eid")
head(covar_dat)
nrow(covar_dat)
```

---

## MACE data from hospital and death records

### Ocurrences in hospital records

Hospital records - main table:

```{r}
hesin <- read_tsv(
    "~/projects/DVA/Data/UKBiobank/ukb_57232/hes_data/hesin.txt",
    guess_max = 50000, show_col_types = FALSE
)
print(data.frame(head(hesin)))
```

Hospital records - Diagnosis table:

```{r}
hesin_diag <- read_tsv(
    "~/projects/DVA/Data/UKBiobank/ukb_57232/hes_data/hesin_diag.txt", 
    skip = 1,
    col_types = "nnn-c-c-",
    col_names = c("eid", "ins_index", "arr_index", "diag_icd9", "diag_icd10")
)
head(hesin_diag)
```

Joining tables:

```{r}
hesin_join <- hesin %>%
    transmute(eid, ins_index, 
              event_date = coalesce(epistart, epiend, elecdate, admidate, epiend, disdate),
              event_date = lubridate::dmy(event_date)) %>%
    inner_join(hesin_diag) %>%
    select(-c(ins_index, arr_index)) %>%
    rename(ICD_9 = diag_icd9, ICD_10 = diag_icd10) %>%
    pivot_longer(c(ICD_9, ICD_10), names_to = "CODESYS", values_to = "CODE", values_drop_na = TRUE)
head(hesin_join)
```

### Occurrences from mortality data

Mortality data - date:

```{r}
mordate <- read_tsv("~/projects/DVA/Data/UKBiobank/ukb_57232/mortality/death.txt", show_col_types = FALSE)
head(mordate)
```

Mortality data - cause:

```{r}
morcause <- read_tsv("~/projects/DVA/Data/UKBiobank/ukb_57232/mortality/death_cause.txt", show_col_types = FALSE)
head(morcause)
```

Mortality - Join tables:

```{r}
mor <- inner_join(mordate, morcause) %>%
    transmute(eid, 
              event_date = lubridate::dmy(date_of_death),
              CODESYS = "ICD_10", CODE = cause_icd10)
head(mor)
```

### Joining occurrences

```{r}
hesmort_outcomes <- bind_rows(
    hesin_join, mor
)
head(hesmort_outcomes)
```

### Finding MACE events

MACE codes:

```{r}
mace_codes <- read_tsv("../data/MACE_CODES.tsv", show_col_types = FALSE)
head(mace_codes)
```

```{r}
outcome_lookup <- mace_codes %>%
    group_by(OUTCOME, CODESYS) %>%
    nest %>%
    mutate(
        data = map_chr(data, ~paste0("^", .x$CODE, collapse = "|"))
    ) %>%
    ungroup
outcome_lookup
```

```{r}
outcome_lookup_res <- outcome_lookup %>%
    transmute(
        outcome = OUTCOME,
        res = map2(
            CODESYS, data,
            ~hesmort_outcomes %>% 
                filter(CODESYS == .x, 
                       grepl(.y, CODE)) %>%
                select(eid, event_date)
        )
    ) %>%
    unnest(res)
head(outcome_lookup_res)
```

### Finding first occurrences of MACE events

```{r}
mace_firstoccur <- outcome_lookup_res %>%
    select(-outcome) %>%
    group_by(eid) %>%
    slice_min(event_date, with_ties = FALSE) %>%
    ungroup
head(mace_firstoccur)
```

### First and last date for all individuals - up to 10 years of follow-up

```{r}
censordate <- lubridate::ymd("2018-02-28")
```

```{r}
datedat <- mor %>%
    select(eid, lastdate = event_date) %>%
    group_by(eid) %>%
    slice_min(lastdate, with_ties = FALSE) %>%
    ungroup %>%
    right_join(crude_dat[,c("eid", "date0")]) %>%
    transmute(
        eid, date0, 
        lastdate = replace_na(lastdate, censordate),
        lastdate = if_else(lastdate > censordate, censordate, lastdate),
        lastdate = if_else(lastdate > date0 + lubridate::years(10), 
                           date0 + lubridate::years(10), lastdate)
    ) %>%
    drop_na
head(datedat)
nrow(datedat)
```

### Preparing data for survival analysis

```{r}
survmacedat <- mace_firstoccur %>%
    transmute(
        eid,
        outcome_value = 1, 
        outcome_date = event_date
    ) %>%
    right_join(datedat, by = "eid") %>%
    filter(eid %in% recoded_dat$eid) %>%
    mutate(
        outcome_value = replace_na(outcome_value, 0),
        outcome_date = coalesce(outcome_date, censordate),
        event_after_censor = outcome_value == 1 & outcome_date > lastdate,
        outcome_value = if_else(event_after_censor, 0, outcome_value),
        outcome_date = if_else(event_after_censor, lastdate, outcome_date),
        outcome_timeyrs = as.numeric(outcome_date - date0) / 365.25
    ) %>%
    filter(outcome_date > date0) %>%
    select(-c(date0, outcome_date, lastdate, event_after_censor))
head(survmacedat)
nrow(survmacedat)
```

---

## Diabetes progression

We will extract information contained in primary care data to assess who developed diabetes after recruitment. The first thing needed is the set of individuals for which primary care data is available - not everyone has data available:

```{r}
system(
    paste(
        "cut", "-f", 1, 
        "/ludc/Home/daniel_c/projects/DVA/Data/UKBiobank/ukb_57232/gp_data/gp_clinical.txt", "|",
        "tail", "-n", "+2", "|",
        "sort", "|",
        "uniq", ">", "../data/primcareinds.tsv"
    )
)
```

```{r}
primcareinds <- read_tsv("../data/primcareinds.tsv", col_names = "eid", col_types = "n")
head(primcareinds)
```

```{r}
nrow(primcareinds)
```

We will only use data from individuals that we include in our primary analysis:

```{r}
primcareinds <- primcareinds %>%
    filter(eid %in% recoded_dat$eid)
nrow(primcareinds)
```

We can add to this set of individuals their corresponding date of recruitment in UK Biobank and last date of follow-up, which we will need for the survival analysis:

```{r}
primcareinds <- primcareinds %>%
    inner_join(datedat, by = "eid")
head(primcareinds)
nrow(primcareinds)
```

We further need to exclude individuals who already reported having diabetes at recruitment:

```{r}
primcareinds <- primcareinds %>%
    inner_join(selected_dx[,c("eid", "T1D", "T2D")], by = "eid") %>%
    inner_join(selected_meds[,c("eid", "Insulin", "AntiDM")])
head(primcareinds)
nrow(primcareinds)
```

```{r}
primcareinds <- primcareinds %>%
    filter(T1D == 0 & T2D == 0 & Insulin == 0 & AntiDM == 0) %>%
    select(-c(T1D, T2D, Insulin, AntiDM))
head(primcareinds)
nrow(primcareinds)
```

Next, we'll extract codes registered in primary care data corresponding to clinical entities equivalent to diabetes were selected from a curated list found [here](https://github.com/drkgyoung/UK_Biobank_codelists). They are organized in the following table:

```{r}
primcarecodes_dm <- read_tsv("../data/PrimaryCareCodesDM.txt", show_col_types = FALSE)
head(primcarecodes_dm)
```

Saving only the codes to look for them in the primary care table:

```{r}
primcarecodes_dm %>%
    select(Code) %>%
    write_tsv("../data/PrimCareCodesDM.tsv", col_names = FALSE)
```

Looking for codes in primary care data table:

```{r}
system(
    paste(
        "grep", "-Fwf", "../data/PrimCareCodesDM.tsv",
        "/ludc/Home/daniel_c/projects/DVA/Data/UKBiobank/ukb_57232/gp_data/gp_clinical.txt", ">",
        "../data/DM_events_PrimCare.tsv"
    )
)
```

Importing results:

```{r}
dmeventsprimcare <- read_tsv(
    "../data/DM_events_PrimCare.tsv", 
    col_names = c("eid", "data_provider", "event_dt", "read_2", "read_3", "value_1", "value_2", "value_3"),
    col_types = "nncccccc"
)
head(dmeventsprimcare)
```

Only individuals included in our analysis:

```{r}
dmevents <- dmeventsprimcare %>%
    transmute(eid, outcome_date = lubridate::dmy(event_dt)) %>%
    inner_join(primcareinds)
head(dmevents)
```

We also need to extract events from hospital records and mortality data. The codes to use are the following:

```{r}
dm_codes_hes <- tribble(
    ~CODESYS, ~CODE,
    "ICD_9", "250",
    "ICD_10", "E10",
    "ICD_10", "E11"
)
```

Similar to what we did for MACE:

```{r}
hesmort_dm <- hesmort_outcomes %>%
    inner_join(primcareinds)
head(hesmort_dm)
length(unique(hesmort_dm$eid))
```

Combining data from the two sources:

```{r}
dmevents <- dm_codes_hes %>%
    group_by(CODESYS) %>%
    nest %>%
    mutate(
        data = map_chr(data, ~paste0("^", .x$CODE, collapse = "|")),
        data = map2(
            CODESYS, data,
            ~hesmort_dm %>% 
                filter(CODESYS == .x, 
                       grepl(.y, CODE)) %>%
                select(eid, outcome_date = event_date, date0, lastdate)
        )
    ) %>%
    unnest(data) %>%
    ungroup %>%
    select(-CODESYS) %>%
    bind_rows(dmevents)
head(dmevents)
```

Now we need the first event recorded:

```{r}
dmevents <- dmevents %>%
    group_by(eid) %>%
    slice_min(outcome_date, with_ties = FALSE) %>%
    ungroup
head(dmevents)
nrow(dmevents)
```

Now formatting into the survival format:

```{r}
survdmdat <- dmevents %>%
    transmute(eid, outcome_value = 1, outcome_date) %>%
    right_join(primcareinds, by = "eid") %>%
    mutate(
        outcome_value = replace_na(outcome_value, 0),
        outcome_date = coalesce(outcome_date, censordate),
        event_after_censor = outcome_value == 1 & outcome_date > lastdate,
        outcome_value = if_else(event_after_censor, 0, outcome_value),
        outcome_date = if_else(event_after_censor, lastdate, outcome_date),
        outcome_timeyrs = as.numeric(outcome_date - date0) / 365.25
    ) %>%
    filter(outcome_date > date0) %>%
    select(-c(date0, outcome_date, lastdate, event_after_censor))
head(survdmdat)
nrow(survdmdat)
```

---

## Saving tables generated

Initial input:

```{r}
write_tsv(recoded_dat, "../data/recoded_dat.tsv")
```

~~~~~~~~~~~~~~~~~~~~~~~

Pre-existing conditions and medication data:

```{r}
write_tsv(covar_dat,  "../data/covar_dat.tsv")
```

~~~~~~~~~~~~~~~~~~~~~~~

Outcome data:

- MACE

```{r}
write_tsv(survmacedat, "../data/survmacedat.tsv")
```

- Time to diabetes diagnosis:

```{r}
write_tsv(survdmdat, "../data/survdmdat.tsv")
```

