---
title: An R Markdown document converted from "11_DataEthnic.ipynb"
output: html_document
---

## Data from other ethnicities

```{r}
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(purrr)
```

---

## Baseline data

```{r}
tab_name <- "~/projects/DVA/Data/UKBiobank/ukb_57232/data_files/41806.tab"
```

Locating columns needed:

```{r}
tab_vars <- tribble(
    ~colid, ~colnam, ~coltyp,
    ## Basic
    "f.eid", "eid", "n",
    "f.53.0.0", "date0", "D",
    "f.31.0.0", "sex", "n",
    "f.21003.0.0", "age", "n",
    ## Anthropometric
    "f.21001.0.0", "bmi", "n",
    "f.48.0.0", "waist_circumference", "n",
    "f.49.0.0", "hip_circumference", "n",
    "f.50.0.0", "height", "n",
    ## Blood pressure
    "f.4080.0.0", "sbp_01", "n",
    "f.4080.0.1", "sbp_02", "n",
    "f.4079.0.0", "dbp_01", "n",
    "f.4079.0.1", "dbp_02", "n",
    ## Liver function
    "f.30620.0.0", "alt", "n",
    ## Kidney function
    "f.30700.0.0", "creatinine", "n",
    ## Inflammatory markers
    "f.30710.0.0", "crp", "n",
    ## Lipid fractions
    "f.30760.0.0", "hdl", "n",
    "f.30780.0.0", "ldl", "n",
    "f.30870.0.0", "tg", "n",
    ## Glycemia
    "f.30740.0.0", "glucose", "n",
    ## Smoking
    "f.1239.0.0", "current_smoker", "n",
    ## Caucasian ancestry
    "f.22006.0.0", "caucasian", "n",
    ## Self-reported ethnicity
    "f.21000.0.0", "srethnicity", "n"
)
```

Collecting positions of these columns in the data:

```{r}
tab_vars <- paste("head -1", tab_name, "| tr '\t' '\n'") %>%
    pipe %>%
    read_table(col_names = "colid", col_types = "c") %>%
    ## Position of columns in table
    mutate(colpos = row_number()) %>%
    ## Extracting position and name of variables of interest
    inner_join(tab_vars, by = "colid")
head(tab_vars)
```

Fasting time (preprocessed from a second application):

```{r}
fast_time <- read_tsv("~/projects/DVA/Data/UKBiobank/fasting_time_ukb.tsv", show_col_types = FALSE)
head(fast_time)
```

Preparing to join with rest of columns:

```{r}
fast_time <- transmute(fast_time, eid = eid_pascal, fasting_time)
head(fast_time)
```

Preparing the `bash` command to extract columns of interest:

```{r}
column_cmd <- paste("cut -f", paste(tab_vars$colpos, collapse = ","), tab_name)
cat(column_cmd)
```

Executing the command and passing to `R`:

```{r}
crude_dat <- read_tsv(
    file = pipe(column_cmd),
    ## Replacing column header with variable names
    skip = 1, col_names = tab_vars$colnam,
    ## All columns are numeric
    col_types = paste(tab_vars$coltyp, collapse = "")
) %>%
    ## Adding fasting time
    inner_join(fast_time, by = "eid")
head(crude_dat)
```

```{r}
nrow(crude_dat)
```

```{r}
recoded_ethn <- crude_dat %>%
    ## Only ancestries other than Caucasian
    filter(is.na(caucasian)) %>%
    select(-caucasian)
nrow(recoded_ethn)
```

```{r}
recoded_ethn$srethnicity %>% table
```

```{r}
ethnicitytab <- tribble(
    ~Ethnicity, ~srethnicity,
    "SouthAsian", c(3001, 3002, 3003),
    "African", c(4, 4001, 4002, 4003),
    "Chinese", c(5)
) %>%
    unnest(srethnicity)
ethnicitytab
```

```{r}
recoded_ethn <- recoded_ethn %>%
    inner_join(ethnicitytab, by = "srethnicity")
nrow(recoded_ethn)
```

```{r}
recoded_ethn <- recoded_ethn %>%
    transmute(eid, age, Ethnicity,
              sex = ifelse(sex == 0, "Female", "Male"),
              ## Anthropometric
              bmi, 
              whr = waist_circumference / hip_circumference,
              ## Blood pressure
              sbp = rowMeans(cbind(sbp_01, sbp_02), na.rm = TRUE),
              dbp = rowMeans(cbind(dbp_01, dbp_02), na.rm = TRUE),
              ## Liver function
              alt,
              ## Kidney function
              scr = creatinine,
              ## Inflammatory markers
              crp,
              ## Lipid fractions
              hdl, tg, ldl,
              ## Glycemia - At least fasting time of 4 hours
              fg = ifelse(fasting_time >= 4, glucose, NaN),
              ## Current smoking status
              smoking = case_when(current_smoker %in% c(1,2) ~ 1,
                                  current_smoker == 0 ~ 0,
                                  TRUE ~ NaN)) %>%
    ## Only complete cases
    drop_na
head(recoded_ethn)
nrow(recoded_ethn)
```

```{r}
table(recoded_ethn$Ethnicity)
```

```{r}
recoded_ethn <- recoded_ethn %>%
    group_by(Ethnicity) %>%
    filter(n() >= 1000) %>%
    ungroup
table(recoded_ethn$Ethnicity)
```

## Covariate data

```{r}
covar_dat <- read_tsv("../data/covar_dat.tsv", show_col_types = FALSE)
head(covar_dat)
```

## MACE data from hospital and death records

### Ocurrences in hospital records

Hospital records - main table:

```{r}
hesin <- read_tsv(
    "~/projects/DVA/Data/UKBiobank/ukb_57232/hes_data/hesin.txt",
    guess_max = 50000, show_col_types = FALSE
)
print(data.frame(head(hesin)))
```

Hospital records - Diagnosis table:

```{r}
hesin_diag <- read_tsv(
    "~/projects/DVA/Data/UKBiobank/ukb_57232/hes_data/hesin_diag.txt", 
    skip = 1,
    col_types = "nnn-c-c-",
    col_names = c("eid", "ins_index", "arr_index", "diag_icd9", "diag_icd10")
)
head(hesin_diag)
```

Joining tables:

```{r}
hesin_join <- hesin %>%
    transmute(eid, ins_index, 
              event_date = coalesce(epistart, epiend, elecdate, admidate, epiend, disdate),
              event_date = lubridate::dmy(event_date)) %>%
    inner_join(hesin_diag) %>%
    select(-c(ins_index, arr_index)) %>%
    rename(ICD_9 = diag_icd9, ICD_10 = diag_icd10) %>%
    pivot_longer(c(ICD_9, ICD_10), names_to = "CODESYS", values_to = "CODE", values_drop_na = TRUE)
head(hesin_join)
```

### Occurrences from mortality data

Mortality data - date:

```{r}
mordate <- read_tsv("~/projects/DVA/Data/UKBiobank/ukb_57232/mortality/death.txt", show_col_types = FALSE)
head(mordate)
```

Mortality data - cause:

```{r}
morcause <- read_tsv("~/projects/DVA/Data/UKBiobank/ukb_57232/mortality/death_cause.txt", show_col_types = FALSE)
head(morcause)
```

Mortality - Join tables:

```{r}
mor <- inner_join(mordate, morcause) %>%
    transmute(eid, 
              event_date = lubridate::dmy(date_of_death),
              CODESYS = "ICD_10", CODE = cause_icd10)
head(mor)
```

### Joining occurrences

```{r}
hesmort_outcomes <- bind_rows(
    hesin_join, mor
)
head(hesmort_outcomes)
```

### Finding MACE events

MACE codes:

```{r}
mace_codes <- read_tsv("../data/MACE_CODES.tsv", show_col_types = FALSE)
head(mace_codes)
```

```{r}
outcome_lookup <- mace_codes %>%
    group_by(OUTCOME, CODESYS) %>%
    nest %>%
    mutate(
        data = map_chr(data, ~paste0("^", .x$CODE, collapse = "|"))
    ) %>%
    ungroup
outcome_lookup
```

```{r}
outcome_lookup_res <- outcome_lookup %>%
    transmute(
        outcome = OUTCOME,
        res = map2(
            CODESYS, data,
            ~hesmort_outcomes %>% 
                filter(CODESYS == .x, 
                       grepl(.y, CODE)) %>%
                select(eid, event_date)
        )
    ) %>%
    unnest(res)
head(outcome_lookup_res)
```

### Finding first occurrences of MACE events

```{r}
mace_firstoccur <- outcome_lookup_res %>%
    select(-outcome) %>%
    group_by(eid) %>%
    slice_min(event_date, with_ties = FALSE) %>%
    ungroup
head(mace_firstoccur)
```

### First and last date for all individuals - up to 10 years of follow-up

```{r}
censordate <- lubridate::ymd("2018-02-28")
```

```{r}
datedat <- mor %>%
    select(eid, lastdate = event_date) %>%
    group_by(eid) %>%
    slice_min(lastdate, with_ties = FALSE) %>%
    ungroup %>%
    right_join(crude_dat[,c("eid", "date0")]) %>%
    transmute(
        eid, date0, 
        lastdate = replace_na(lastdate, censordate),
        lastdate = if_else(lastdate > censordate, censordate, lastdate),
        lastdate = if_else(lastdate > date0 + lubridate::years(10), 
                           date0 + lubridate::years(10), lastdate)
    ) %>%
    drop_na
head(datedat)
nrow(datedat)
```

### Preparing data for survival analysis

```{r}
survmacedat <- mace_firstoccur %>%
    transmute(
        eid,
        outcome_value = 1, 
        outcome_date = event_date
    ) %>%
    right_join(datedat, by = "eid") %>%
    filter(eid %in% recoded_ethn$eid) %>%
    mutate(
        outcome_value = replace_na(outcome_value, 0),
        outcome_date = coalesce(outcome_date, censordate),
        event_after_censor = outcome_value == 1 & outcome_date > lastdate,
        outcome_value = if_else(event_after_censor, 0, outcome_value),
        outcome_date = if_else(event_after_censor, lastdate, outcome_date),
        outcome_timeyrs = as.numeric(outcome_date - date0) / 365.25
    ) %>%
    filter(outcome_date > date0) %>%
    select(-c(date0, outcome_date, lastdate, event_after_censor))
head(survmacedat)
nrow(survmacedat)
```

---

## Diabetes progression

```{r}
primcareinds <- read_tsv("../data/primcareinds.tsv", col_names = "eid", col_types = "n")
head(primcareinds)
```

```{r}
nrow(primcareinds)
```

We will only use data from individuals that we include in our primary analysis:

```{r}
primcareinds <- primcareinds %>%
    filter(eid %in% recoded_ethn$eid)
nrow(primcareinds)
```

We can add to this set of individuals their corresponding date of recruitment in UK Biobank and last date of follow-up, which we will need for the survival analysis:

```{r}
primcareinds <- primcareinds %>%
    inner_join(datedat, by = "eid")
head(primcareinds)
nrow(primcareinds)
```

We further need to exclude individuals who already reported having diabetes at recruitment:

```{r}
primcareinds <- primcareinds %>%
    inner_join(covar_dat[,c("eid", "T1D", "T2D", "Insulin", "AntiDM")], by = "eid")
head(primcareinds)
nrow(primcareinds)
```

```{r}
primcareinds <- primcareinds %>%
    filter(T1D == 0 & T2D == 0 & Insulin == 0 & AntiDM == 0) %>%
    select(-c(T1D, T2D, Insulin, AntiDM))
head(primcareinds)
nrow(primcareinds)
```

Next, we'll extract codes registered in primary care data corresponding to clinical entities equivalent to diabetes were selected from a curated list found [here](https://github.com/drkgyoung/UK_Biobank_codelists). They are organized in the following table:

Importing results:

```{r}
dmeventsprimcare <- read_tsv(
    "../data/DM_events_PrimCare.tsv", 
    col_names = c("eid", "data_provider", "event_dt", "read_2", "read_3", "value_1", "value_2", "value_3"),
    col_types = "nncccccc"
)
head(dmeventsprimcare)
```

Only individuals included in our analysis:

```{r}
dmevents <- dmeventsprimcare %>%
    transmute(eid, outcome_date = lubridate::dmy(event_dt)) %>%
    inner_join(primcareinds)
head(dmevents)
```

We also need to extract events from hospital records and mortality data. The codes to use are the following:

```{r}
dm_codes_hes <- tribble(
    ~CODESYS, ~CODE,
    "ICD_9", "250",
    "ICD_10", "E10",
    "ICD_10", "E11"
)
```

Similar to what we did for MACE:

```{r}
hesmort_dm <- hesmort_outcomes %>%
    inner_join(primcareinds)
head(hesmort_dm)
length(unique(hesmort_dm$eid))
```

Combining data from the two sources:

```{r}
dmevents <- dm_codes_hes %>%
    group_by(CODESYS) %>%
    nest %>%
    mutate(
        data = map_chr(data, ~paste0("^", .x$CODE, collapse = "|")),
        data = map2(
            CODESYS, data,
            ~hesmort_dm %>% 
                filter(CODESYS == .x, 
                       grepl(.y, CODE)) %>%
                select(eid, outcome_date = event_date, date0, lastdate)
        )
    ) %>%
    unnest(data) %>%
    ungroup %>%
    select(-CODESYS) %>%
    bind_rows(dmevents)
head(dmevents)
```

Now we need the first event recorded:

```{r}
dmevents <- dmevents %>%
    group_by(eid) %>%
    slice_min(outcome_date, with_ties = FALSE) %>%
    ungroup
head(dmevents)
nrow(dmevents)
```

Now formatting into the survival format:

```{r}
survdmdat <- dmevents %>%
    transmute(eid, outcome_value = 1, outcome_date) %>%
    right_join(primcareinds, by = "eid") %>%
    mutate(
        outcome_value = replace_na(outcome_value, 0),
        outcome_date = coalesce(outcome_date, censordate),
        event_after_censor = outcome_value == 1 & outcome_date > lastdate,
        outcome_value = if_else(event_after_censor, 0, outcome_value),
        outcome_date = if_else(event_after_censor, lastdate, outcome_date),
        outcome_timeyrs = as.numeric(outcome_date - date0) / 365.25
    ) %>%
    filter(outcome_date > date0) %>%
    select(-c(date0, outcome_date, lastdate, event_after_censor))
head(survdmdat)
nrow(survdmdat)
```

---

## Saving tables generated

Initial input:

```{r}
write_tsv(recoded_ethn, "../data/recoded_ethn.tsv")
```

Outcome data:

- MACE

```{r}
write_tsv(survmacedat, "../data/survmacedat_ethn.tsv")
```

- Time to diabetes diagnosis:

```{r}
write_tsv(survdmdat, "../data/survdmdat_ethn.tsv")
```

