---
title: An R Markdown document converted from "19_LassoCox.ipynb"
output: html_document
---

# Running Lasso

---

## Libraries

```{r}
library(readr)
library(dplyr, warn.conflicts = FALSE)
library(tidyr, warn.conflicts = FALSE)
library(purrr)
library(survival)
library(Matrix, warn.conflicts = FALSE)
library(glmnet, quietly = TRUE)
library(ggplot2)
```

```{r}
options(repr.plot.res = 300)
```

## Pre-defined functions

```{r}
source("18_LassoCox.R")
```

## Stratified data

```{r}
load("../data/ukb/strat_dat.RData")
```

```{r}
map(strat_dat, head)
```

## Validated clusters

```{r}
load("../data/validclusmod.RData")
print(validclusmod)
```

## Pre-existing conditions and medications

```{r}
covar_dat <- read_tsv("../data/covar_dat.tsv", show_col_types = FALSE)
head(covar_dat)
```

## Survival data

### MACE

```{r}
survmacedat <- read_tsv("../data/survmacedat.tsv", show_col_types = FALSE)
head(survmacedat)
```

### DM

```{r}
survdmdat <- read_tsv("../data/survdmdat.tsv", show_col_types = FALSE)
head(survdmdat)
```

## Calculation of cluster probabilities

```{r}
clusterdfs <- clusterprobcalc(ClusModDf = validclusmod, StratDat = strat_dat)
```

```{r}
print(clusterdfs)
```

## Adding covariates

```{r}
clusterdfs <- addcovardat(X = clusterdfs, CovarDat = covar_dat)
```

```{r}
print(clusterdfs)
```

## Adding survival data

```{r}
clustersurvdfs <- addsurvdat(X = clusterdfs, SurvMACEDf = survmacedat, SurvDMDf = survdmdat)
```

```{r}
print(clustersurvdfs)
```

## Creating follow-up subsets

```{r}
clustersurvdfs <- futsubsetsfx(clustersurvdfs)
print(clustersurvdfs)
```

## Running Cox models

```{r}
coxmoddf <- coxmodels(clustersurvdfs)
```

```{r}
print(coxmoddf)
```

## Extracting penalized estimates

```{r}
bestlasso <- coxmoddf %>%
    transmute(
        sex, outcome, fut,
        mod_clus = map(
            mod_clus, 
            ~data.frame(Cluster = rownames(.x), estimate = .x[,1])
        )
    ) %>%
    unnest(mod_clus) %>%
    filter(grepl("^alr", Cluster)) %>%
    mutate(Cluster = gsub("alrprob", "", Cluster))
head(bestlasso)
```

## Estimates from standard models

```{r}
coxres <- rio::import("../data/ukb/result_file2.RData") %>%
    pluck("SurvCoefs") %>%
    filter(model == "clus") %>%
    transmute(
        sex, outcome, fut,
        estimates = map2(
            estimates, varcovmat,
            ~data.frame(
                Cluster = names(.x),
                estimate = .x,
                se = sqrt(diag(.y))
            )
        )
    ) %>%
    unnest(estimates) %>%
    filter(grepl("^alr", Cluster)) %>%
    mutate(Cluster = gsub("alrprob", "", Cluster))
head(coxres)
```

## Cluster map

```{r}
cluscolmap <- validclusmod %>%
    transmute(
        clusmod = map(clusmod, select, Cluster = validclus_name, W = validclus_weights)
    ) %>%
    unnest(clusmod) %>%
    group_by(Cluster) %>%
    summarise(W2 = exp(sum(log((W))))) %>%
    arrange(-W2) %>%
    mutate(
        Cluster = factor(Cluster, levels = Cluster),
        ClusCol = scales::hue_pal()(n())
    )
cluscolmap
```

## Comparing standard and Lasso estimates

```{r}
options(repr.plot.width = 5, repr.plot.height = 3)
coxres %>%
    mutate(
        esttyp = "Standard"
    ) %>%
    bind_rows(bestlasso) %>%
    mutate(
        esttyp = replace_na(esttyp, "Lasso"),
        Cluster = factor(Cluster, levels = rev(cluscolmap$Cluster)),
        ci = qnorm(1 - .05/2) * se,
        ci = replace_na(ci, 0),
        lwr = estimate - ci,
        upr = estimate + ci
    ) %>%
    ggplot(aes(estimate, Cluster)) +
    geom_vline(xintercept = 0, lty = "dashed") +
    geom_linerange(
        aes(xmin = lwr, xmax = upr)
    ) +
    geom_point(
        aes(color = esttyp, shape = esttyp)
    ) +
    scale_color_manual(
        name = "Estimate",
        values = c(Standard = "black", Lasso = "red")
    ) +
    scale_shape_manual(
        name = "Estimate",
        values = c(Standard = 19, Lasso = 18)
    ) +
    ggh4x::facet_nested(
        sex ~ outcome + fut, 
        scales = "free", space = "free_y",
    ) +
    scale_x_continuous(n.breaks = 4, labels = \(x)round(x, 2)) +
    labs(x = "Estimate\n95% CI", y = NULL) +
    theme_bw() +
    theme(
        axis.text.x = element_text(size = 5),
        axis.text.y = element_text(size = 6)
    )
```

